Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-08-29 paper/letter.tex
Max comments

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ emission reveals a dispersion-supported dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of $69^{+17}_{-15}$ km s$^{-1}$ with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plaussible explanation is provided by the
multiphase model as it is a close match to the galaxy scaling relationship
for a dark matter dominated dwarf elliptical supported by its
velocity dispersion. 
This work demonstrates the importance of considering multiphase
physics among the possible conditions shaping the \lya\ spectra of the
first galaxies.   Additionally, if future kinematic maps of
\tol\ confirm the high velocities postulated in our model, it would
help to stablish the \lya\ line morphology as an observational tracer to
study the co-evolution of the first generation of galaxies and its
host dark matter halos. 
}

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=69^{+17}_{-15}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+73}_{-42}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  $P_{cl}=0.78^{+0.16}_{-0.43}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$. In the rotation
model assuming reasonable bounds for the number density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

@@

A future observational test to clarify the physical nature of
\tol\ would require integral field unit measurements spatially
resolving its spatial extent. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the escape
fraction of Ly continuum ionizing radiation. 
In the rotational model this fraction should be zero, while
the multiphase model predicts that in this case it should be around
$0.5^{+1.0}_{-0.4}$\%.

@@

%\url{http://vizier.u-strasbg.fr/cgi-bin/VizieR?-source=B/2mass}
% TOLOLO 00021
%https://ned.ipac.caltech.edu/cgi-bin/objsearch?objname=Tol+21&extend=no&hconst=73&omegam=0.27&omegav=0.73&corr_z=1&out_csys=Equatorial&out_equinox=J2000.0&obj_sort=RA+or+Longitude&of=pre_text&zv_breaker=30000.0&list_limit=5&img_stamp=YES 
\tol receding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (Hubble constant 73 Mpc km$^{-1}$
s$^{1}$)
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.

@@

There is a concise analytical description for those features.
This description takes into account how different parts of the
sphere's surface shift in frequency the \lya photons. 
Different shifts in frequency come from different values for the projected
velocity along the LOS. 
As presented in \cite{CLARA}, using the analytical solution for the
\lya sphectra of a static sphere plus the right frequency shifts
computed from geometrical considerations, one is able to produce an
analytical solution for the rotating sphere that reproduces the main
features found using the full numerical simulation.

The analytical solution for the rotation sphere was the base to
perform the Markov Chain Monte Carlo Calculation using the
\texttt{emcee} implementation.
We explore flat priors on $V_{\rm max}$ $\log_{10}\tau$, $\log_{10} T$ and
$\theta$ using $XX$ steps.
The results are summarized in Figure $XX$. 
From this model we find that the fiducial parameters that could
explain the broad features in \tol\ are $V_{\rm max}=348^{+75}_{-48}$
km s$^{-1}$, $\log \tau = 6.96^{+0.26}_{-0.18}$, $\log_{10} T/\mathrm
{K} = 4.27^{+0.11}_{-0.18}$ and  $\theta = 35.78^{+2.13}_{-1.88}$ degrees.

\begin{figure}
\begin{center}
\includegraphics[width=1.0\textwidth]{emcee_results.pdf}
\caption{{\bf Results from the Markov Chain Monte Carlo compuation for the rotation model.}}
\end{center}
\end{figure}

@@

The idealized multiphase model consists of spherical, cold, dens
clumps of neutral hydrogen (and dust) embedded in a hot, ionized
medium. 
The clumps also have a random and an outflowing velocity
component which totals the number of parameters describing the model
to be $14$. 
In order to map out this large parameter space, we randomly drew
$2500$ sets of parameters within a observationally realistic range
(based on the considerations of \cite{Laursen2013ApJ...766..124L})
yielding a large variety of single-, double- and triple-peaked
spectra. 
The full analysis of the the spectral features as well as
more details on the radiative transfer are presented in
\cite{Gronke2016}.

For the current work, we computed the $\chi^2$ for each of the $2500$
parameters yielding the best fit parameters of $(...)$. Here, the
subscripts ${}_{\rm cl}$ and ${}_{\rm ICM}$ stand for the quantities
filling the clumps and the medium between the clumps,
respectively. 
Furthermore, we found that some parameters such as the magnitude of
the random clump motion $\sigma_{\rm cl}$ improved the fit
significantly whereas others did not.

@@

From the multiphase model we obtained $v=69_{-15}^{+17}$ \kms
and $r=5$ kpc. 
This corresponds to a dynamical mass of $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$ which is at least
$5$ times the HI mass plus the stellar mass estimated from
observations.

>>>>

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\sigmaclump}{$69^{+17}_{-15}$ km s$^{-1}$}
\newcommand{\inftyclump}{$79^{+73}_{-42}$ km s$^{-1}$}
\newcommand{\probaclump}{$0.78^{+0.16}_{-0.43}$}
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ emission reveals a dispersion-supported dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plaussible explanation is provided by the
multiphase model as it is a close match to the galaxy scaling relationship
for a dark matter dominated dwarf elliptical supported by its
velocity dispersion. 
This work demonstrates the importance of considering multiphase
physics among the possible conditions shaping the \lya\ spectra of the
first galaxies.   Additionally, if future kinematic maps of
\tol\ confirm the high velocities postulated in our model, it would
help to stablish the \lya\ line morphology as an observational tracer to
study the co-evolution of the first generation of galaxies and its
host dark matter halos. 
}

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=$\sigmaclump\
the clumps outflowing velocity $v_{\infty, \rm{cl}}=$\inftyclump\
and the fraction of the \lya\ emission that is  coming
from the cold clumps  $P_{\rm cl}=$\probaclump.
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$. In the rotation
model assuming reasonable bounds for the number density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

@@

A future observational test to clarify the physical nature of
\tol\ would require integral field unit measurements spatially
resolving its spatial extent. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the escape
fraction of \lya\ continuum ionizing radiation. 
In the rotational model this fraction should be zero, while
the multiphase model predicts that in this case it should be around
$0.5^{+1.0}_{-0.4}$\%.

@@

%\url{http://vizier.u-strasbg.fr/cgi-bin/VizieR?-source=B/2mass}
% TOLOLO 00021
%https://ned.ipac.caltech.edu/cgi-bin/objsearch?objname=Tol+21&extend=no&hconst=73&omegam=0.27&omegav=0.73&corr_z=1&out_csys=Equatorial&out_equinox=J2000.0&obj_sort=RA+or+Longitude&of=pre_text&zv_breaker=30000.0&list_limit=5&img_stamp=YES 
\tol receding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (with the Hubble constant $H_{0}$=73
Mpc km s$^{-1}$) 
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.

@@

There is a concise analytical description for those features.
This description takes into account how different parts of the
sphere's surface shift in frequency the \lya photons. 
Different shifts in frequency come from different values for the projected
velocity along the LOS. 
As presented in \cite{CLARA}, using the analytical solution for the
\lya\ spectra of a static sphere plus the right frequency shifts
computed from geometrical considerations, one is able to produce an
analytical solution for the rotating sphere that reproduces the main
features found using the full numerical simulation.

The analytical solution for the rotation sphere was the base to
perform the Markov Chain Monte Carlo Calculation using the
\texttt{emcee} implementation.
We explore flat priors on $V_{\rm max}$ $\log_{10}\tau$, $\log_{10} T$ and
$\theta$ using $XX$ steps.
The results are summarized in Figure \ref{emceeresults}. 
From this model we find that the fiducial parameters that could
explain the broad features in \tol\ are $V_{\rm max}=348^{+75}_{-48}$
km s$^{-1}$, $\log \tau = 6.96^{+0.26}_{-0.18}$, $\log_{10} T/\mathrm
{K} = 4.27^{+0.11}_{-0.18}$ and  $\theta = 35.78^{+2.13}_{-1.88}$ degrees.

\begin{figure}
\begin{center}
\includegraphics[width=1.0\textwidth]{emcee_results.pdf}
\caption{{\bf Results from the Markov Chain Monte Carlo compuation for
    the rotation model.}\label{emceeresults}} 
\end{center}
\end{figure}

@@

The idealized multiphase model consists of spherical, cold, dens
clumps of neutral hydrogen (and dust) embedded in a hot, ionized
medium. 
The clumps also have a random and an outflowing velocity
component which totals the number of parameters describing the model
to be $14$.

In order to map out this large parameter space, we randomly drew
$2500$ sets of parameters within a observationally realistic range
(based on the considerations of \cite{Laursen2013ApJ...766..124L})
yielding a large variety of single-, double- and triple-peaked
spectra. 
The full analysis of the the spectral features as well as
more details on the radiative transfer are presented in
\cite{Gronke2016}.

For the current work, we computed the $\chi^2$ for each of the $2500$
Furthermore, we found that some parameters such as the magnitude of
the random clump motion $\sigma_{\rm cl}$ improved the fit
significantly whereas others did not.

We find that the best constrained parameters are the clump velocity
dispersion $\sigma_{\rm cl}=$\sigmaclump,  the outflowing clump
velocity $v_{\infty {\rm, cl}}=$\inftyclump\ and the probability that
the \lya\ emission comes from the clumps $P_{\rm cl}=$\probaclump.

@@

From the multiphase model we obtained $v=$\sigmaclump
and $r=5$ kpc. This corresponds to a dynamical mass of $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$ which is at least
$5$ times the HI mass plus the stellar mass estimated from
observations.
